---
title: "Task_2.2"
output: html_document
date: "2025-11-19"
---

# Task 2.2
## PCA Group Work

Utilizing the dataset from the paper (), you will work in small groups to run a PCA and create a scree plot and biplot to visualize the results. Use the below code as a reference if need be or work with python if you are more familiar with python libraries for this analysis. We will go over group results briefly.

### Information on the data being used for this analysis

**Understanding the data:**

* **Metadata / categorical columns:**
  * Mouse replicate
  * Mouse
  * timepoint after AB application
  * Day
  * Group

* **Quantitative measurements:**

  * Raw abundances: KB1, YL2, etc.
  * Log-transformed values: log10 KB1, log10 YL2, etc.

PCA only works on numeric data, so we’ll focus on the numeric columns (either the raw counts or the log-transformed ones — usually log-transformed is better for microbiome or abundance data to reduce skew).

### 1. Preparing Data:

```{r}
library(dplyr)

df_mouse <- read.csv("/second_day_part1/data/mmc2.csv", stringsAsFactors = FALSE)
head(df_mouse)
```

Remove NA values and then take only the log 10 values with the mouse and day metadata. We use the log10 abundance values in this case.

```{r}
# Remove the NA values from the dataframe:
df_na_removed <- na.omit(df_mouse) # This will be used as metadata later on

# Keep only log columns + mouse + day
df_log <- df_na_removed %>% select(Mouse, Day, starts_with("log10"))
head(df_log)
```

Now we want to make the dataset per mouse with abundance values per day. We use the tidyr package a part of `dpylr`.

```{r}
wide <- tidyr::pivot_wider(
  df_log,
  names_from = Day,
  values_from = starts_with("log10"),
  names_glue = "{.value}_Day{Day}"
)

head(wide)
```

Note we do see some NA values here because the 1682 mouse only has values from day 0 to day 9, after that it has no measurements. Thus when we rotate the dataframe we get missing values for those reads.

```{r}
# Check to see if there is the same number of mice once the table has been pivoted:
cat("Number of Mice pre-pivot: ")
length(unique(df_log$Mouse)) # We take the number `length` of the unique Mouse IDs in the column "Mouse"

cat("Number of mice post-pivot: ")
length(unique(wide$Mouse))
```

Next we want to get the day 0 to day 9 difference in abundance values. Of course you can look at other time differences as well.

Here's a breakdown of the code:

1. `mutate(across(...))`
* mutate() creates new columns.
* across() tells mutate: apply the same operation to multiple columns.

2. `contains("Day9")`
* This selects every column whose name contains "Day9".

3. The function: `~ . - get(sub("_Day9","_Day0", cur_column()))`

* cur_column() - Returns the current column name, e.g. "KB1_Day9".
* sub("_Day9","_Day0", cur_column()) - Replaces "Day9" with "Day0".
  * "KB1_Day9" → "KB1_Day0"
  * "YL2_Day9" → "YL2_Day0"

* get(...) - Retrieves that column from the dataframe.
* The .~ means:
  * . = the value of the current Day9 column
  * get(...) = the matching Day0 column

**So the expression does: Day9_value - Day0_value**

4. `.names = "{.col}_delta9"`
* This tells mutate() how to name the new column.

```{r}
delta_df <- wide %>%
  mutate(across(contains("Day9"),
                ~ . - get(sub("_Day9","_Day0", cur_column())),
                .names = "{.col}_delta9"))

head(delta_df)
```

```{r}
# Keeping mouse column for metadata but selecting the columns containing the difference in abundances
day9_delta <- delta_df %>% select(Mouse, contains("delta"))
#day9_delta <- na.omit(day9_delta)

# Now creating a separate dataframe for the numerical only values to be used in the PCA
day9_delta_only <- day9_delta %>% select(contains("delta"))

head(day9_delta_only)
```

### Running the PCA:

Here we use the `prcomp` function to run a PCA:

```{r}
# Running the PCA
pca_result_delta <- prcomp(day9_delta_only, center = TRUE, scale = FALSE)
summary(pca_result_delta)
```

```{r}
pca_result_delta$rotation   # loadings (eigen vectors)
```

### Visualizations:

Load in the necessary ggplot libraries:

```{r}
library(ggplot2)
library(ggfortify)
```

#### Scree Plot
Use ggplot to create a scree plot of the variance for each principal component.

```{r}
library(ggplot2)

# Getting the variance explained by each PC - the same way we did in the examples
eig_vals_pulse <- pca_result_delta$sdev^2
var_explained_pulse <- eig_vals_pulse / sum(eig_vals_pulse)

# Looking at the variance:
cat("Variance explained: ")
var_explained_pulse
```

```{r}
# Plotting the scree plot for PCs
qplot(c(1:13), var_explained_pulse) + # Here we input the data to be plotted x = vector(1,2,3,4), y = the variance
  geom_line() + # We specify that we want a line plot
  xlab("Principal Component") + # Specify the x labels
  ylab("Variance Explained") + # Specify the y axis labels
  ggtitle("Scree Plot") + # Create a title for the plot
  ylim(0, 1) # Show everything from 0 to 1
```

#### Biplot

Use ggplot to create a biplot of the first PC. Same as what we saw in the examples that we did above. The plot below is sufficient, though please play around with some of the features in ggplot to make it more readable all in all (as is seen with the second biplot).

```{r}
# Getting the metadata for coloring in the plot
metadata <- df_na_removed %>% select(Mouse, Group)
metadata_mice <- unique.data.frame(metadata)

# Creating the biplot
autoplot(pca_result_delta,
         data = metadata_mice,
         colour = 'Group',
         loadings = TRUE,
         loadings.label = TRUE,
         loadings.label.size = 3)
```

Here is an example of a more complex biplot, colored by treatment and the mouse point are increased in size as well as arrows changed to black with improved labeling to reduce overlapping. Feel free to change and play around with the settings here to see what best represents the data in the way you want.

```{r}
autoplot(pca_result_delta,
         data = metadata_mice,
         colour = 'Group',
         loadings = TRUE,
         loadings.label = TRUE,   # Add loadings labels
         loadings.label.size = 3, # Adjust label size for readability
         loadings.colour = "black", # Make the loadings labels black for better contrast
         loadings.label.repel = TRUE, # Use repelling labels to avoid overlaps
         size = 4,                # Adjust the size of the points
         alpha = 0.7) +           # Adjust transparency to make points more distinguishable
  labs(title = "PCA Biplot with Day 9 vs Day 0 Abundance Differences",
       x = paste("PC1 (", round(53.6, 1), "%)", sep = ""),
       y = paste("PC2 (", round(21.7, 1), "%)", sep = "")) +
  theme_minimal() +
  theme(legend.position = "right",  # Move the legend to the right
        legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 14)) # Center title
```





