---
title: "Day 2 – 00 Prepare Environment"
author: "Seminar setup"
output:
  html_document:
    toc: false
---

Use this short notebook (or just run the snippets below) before the workshop to
ensure all required packages are installed. Keep it simple: install everything
in one go, then load each package once to check that it works.

```{r install, eval=FALSE}
install.packages(c('ComplexHeatmap', 'circlize', 'tidyr', 'dplyr'))
```

```{r load, eval=FALSE}
library(ComplexHeatmap)
library(circlize)
library(tidyr)
library(dplyr)
```

After installing the R packages, confirm that Pandoc is present. RStudio ships
with Pandoc; otherwise install it manually (e.g., `brew install pandoc` on
macOS or download the installer from <https://pandoc.org/installing.html>). The
Homebrew install places the binary in `/opt/homebrew/bin/pandoc`, which R picks
up automatically.

```{r check-pandoc}
if (!rmarkdown::pandoc_available('1.12.3')) {
  message('Pandoc 1.12.3+ not found. Install it before knitting the Rmd files.')
} else {
  message('Pandoc version detected: ', rmarkdown::pandoc_version())
}
```

If you prefer a one-liner from the terminal, run:

```
R -e "pkgs <- c('ComplexHeatmap','circlize','tidyr','dplyr'); install.packages(setdiff(pkgs, rownames(installed.packages())))"
```

After installing, you can knit `second_day_part2/data/00_prepare_dataset.Rmd` to
create the teaching datasets (if needed) and move on to the exploration scripts.
