---
title: "Day 2 – 01 Explore Data (Exercises – Solutions)"
author: "Seminar practice worksheet"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Use this key after attempting `scripts/01_explore_data_exercises.Rmd`. The code
chunks mirror the TODOs but include one possible solution for each task.

## 1. Load and preview the dataset

```{r ex1-sol}
input_path <- file.path('..', 'data', 'dataset2_subset_long.csv')
long_df <- read.csv(input_path, stringsAsFactors = FALSE, check.names = FALSE)
cat('Rows:', nrow(long_df), '\\nColumns:', ncol(long_df), '\\n')
head(long_df)
```

## 2. Enumerate genomes and SNPs

```{r ex2-sol}
unique(long_df$Genome)

table(long_df$Genome)

tapply(long_df$snp_id, long_df$Genome, function(x) length(unique(x)))
```

## 3. Summaries by genome

```{r ex3-sol}
sample_values <- long_df$value[1:10]
summary(sample_values)

aggregate(value ~ Genome, data = long_df, function(x) summary(x))
```

## 4. Focus on Turicimonas

```{r ex4-sol}
turicimonas_df <- long_df[long_df$Genome == 'Turicimonas_muris_YL45', ]
cat('Rows for Turicimonas:', nrow(turicimonas_df), '\\n')
head(turicimonas_df)

table(turicimonas_df$mouse_id, turicimonas_df$day)
```

## 5. Allele-frequency focus (Turicimonas vs all genomes)

```{r ex5-sol, fig.width=6, fig.height=4}
turicimonas_values <- long_df$value[long_df$Genome == 'Turicimonas_muris_YL45']
hist(turicimonas_values, breaks = 20, main = 'Turicimonas AF', xlab = 'value')
summary(turicimonas_values)
```

```{r ex5b-sol, fig.width=6, fig.height=4}
hist(long_df$value, breaks = 30, main = 'All genomes AF', xlab = 'value')
summary(long_df$value)
```

## 6. Missing values

```{r ex6-na}
na_total <- sum(is.na(long_df$value))
na_total

na_by_mouse_day <- with(long_df, tapply(value, list(mouse_id, day), function(x) sum(is.na(x))))
na_by_mouse_day
```

In this dataset the counts are all zero, so no genome (including Turicimonas)
introduces missing allele-frequency entries.

## 7. Treatment groups

```{r ex7-groups}
unique(long_df$treatment_group)
table(long_df$treatment_group)
table(long_df$mouse_id, long_df$treatment_group)
long_df[long_df$day == 30, c('mouse_id','treatment_group')]
```

## 8. Stretch idea (optional)

```{r ex8-sol}
day30 <- long_df[long_df$day == 30, ]
medians_day30 <- tapply(day30$value, day30$Genome, median, na.rm = TRUE)
medians_day30
medians_day30[which.max(medians_day30)]
```

Feel free to compare your answers with these outputs, then proceed to
`scripts/02_simple_heatmap.Rmd`.
