title: "Day 2 – 01 Explore Data (Exercises)"
author: "Seminar practice worksheet"
output:
  html_document:
    toc: true
    toc_depth: 2
---

This self-paced worksheet mirrors the guided exploration but uses
`second_day_part2/data/dataset2_subset_long.csv`, which contains three genomes
(Akkermansia, Bacteroides, Turicimonas). Students should type their own code
into the chunks below. Hints are provided in the text; the code blocks are left
with `# TODO` comments.

> **Tip:** Run `second_day_part2/00_prepare.Rmd` once to install packages, and
> `second_day_part2/data/00_prepare_dataset.Rmd` only if you need to regenerate
> the three-genome dataset.

## 1. Load and preview the dataset

Goal: use `read.csv()` to load the file, then report the number of rows/columns
and display the first few entries.

```{r ex1, eval=FALSE}
# TODO: load dataset2_subset_long.csv into an object named long_df
# TODO: print nrow(long_df) and ncol(long_df)
# TODO: call head(long_df)
```

*Hint:* remember to set `stringsAsFactors = FALSE` and `check.names = FALSE` so
the column headers stay untouched.

## 2. Enumerate genomes and SNPs

Goal: list all genomes present, count how many rows each contributes, and count
unique SNP identifiers per genome.

```{r ex2, eval=FALSE}
# TODO: unique() to list genomes
# TODO: table() to count rows per genome
# TODO: tapply() + length(unique()) to count SNPs per genome
```

*Hint:* model your solution after the formula `tapply(long_df$snp_id,
long_df$Genome, ...)`.

## 3. Summaries by genome

Goal: compute summary statistics of `value` for each genome (min/median/mean,
etc.) using `aggregate(value ~ Genome, ...)`. Also, show a quick `summary()` of
the first few values to illustrate what the anonymous function does.

```{r ex3, eval=FALSE}
# TODO: sample_values <- long_df$value[1:10]; summary(sample_values)
# TODO: aggregate(value ~ Genome, data = long_df, function(x) summary(x))
```

*Hint:* read `value ~ Genome` as “value grouped by Genome”. The column to the
left of `~` must be numeric.

## 4. Focus on Turicimonas

Goal: subset the data frame to `Turicimonas_muris_YL45` only, show `head()` of
the subset, and build a `table(mouse_id, day)` for that genome.

```{r ex4, eval=FALSE}
# TODO: turicimonas_df <- long_df[long_df$Genome == 'Turicimonas_muris_YL45', ]
# TODO: head(turicimonas_df)
# TODO: table(turicimonas_df$mouse_id, turicimonas_df$day)
```

*Hint:* inspect `head(long_df$Genome == 'Turicimonas_muris_YL45')` to see the
TRUE/FALSE mask created by the comparison.

## 5. Visual checkpoints

Goal: create simple bar plots showing how many measurements exist per day and
per mouse.

```{r ex5, eval=FALSE}
# TODO: day_counts <- table(long_df$day); barplot(day_counts,
#       main = 'Samples per day', xlab = 'Day', ylab = 'Count')
# TODO: mouse_counts <- table(long_df$mouse_id); barplot(mouse_counts,
#       main = 'Samples per mouse', xlab = 'Mouse', ylab = 'Count', las = 2)
```

*Hint:* Discuss which day/mouse dominates the dataset and why.

## 6. Stretch idea (optional)

Can you reuse these skills to answer: “Which genome has the highest median
value on day 30?” Outline your approach below before coding it up.

```{r ex6-notes, eval=FALSE}
# Notes / pseudo-code:
# 1. Filter long_df to day == 30
# 2. Split by Genome and compute median(value)
# 3. Identify the maximum
```

These practice tasks mirror the guided walkthrough but force you to re-create
the analysis yourself. Once satisfied, continue with
`scripts/02_simple_heatmap.Rmd` to generate the heatmap.
