---
title: "Day 2 – 04 Full Heatmap (Exercises Solution)"
author: "Seminar reference solution"
output:
  html_document:
    toc: true
    toc_depth: 2
---

This solution notebook offers one way to complete the full heatmap exercise.
Feel free to tweak palettes, ordering, or filtering thresholds to suit your
teaching needs.

## 1. Packages, paths, helpers

```{r setup, message=FALSE}
suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(circlize)
  library(viridisLite)
})
subset_path <- file.path('..','data','dataset3_subset.csv')
long_path   <- file.path('..','data','dataset3_subset_long.csv')
pdf_path    <- file.path('..','pdf','04_full_heatmap_exercise.pdf')
na_color <- '#dcdcdc'
```

## 2. Load data and NA report

```{r load-data}
wide_df <- read.csv(subset_path, check.names = FALSE, stringsAsFactors = FALSE)
long_df <- read.csv(long_path, check.names = FALSE, stringsAsFactors = FALSE)
cat('Wide rows x cols:', nrow(wide_df), ncol(wide_df), '\n')
cat('Long rows x cols:', nrow(long_df), ncol(long_df), '\n')

na_total <- sum(is.na(long_df$value))
cat('NA count (value):', na_total, '\n')
if (na_total > 0) {
  na_table <- with(long_df, tapply(value, list(mouse_id, day), function(x) sum(is.na(x))))
  print(na_table)
}
```

## 3. Matrix + metadata

```{r matrix}
sample_cols <- setdiff(names(wide_df), c('Genome','snp_id','Position'))
mat <- as.matrix(wide_df[, sample_cols])
mode(mat) <- 'numeric'
rownames(mat) <- paste(wide_df$Genome, wide_df$snp_id, sep = ' | ')

sample_meta <- unique(long_df[, c('mouse_id','day','treatment_group')])
sample_meta$sample_id <- paste(sample_meta$mouse_id, sample_meta$day, sep='-')
sample_meta <- sample_meta[match(colnames(mat), sample_meta$sample_id), ]
stopifnot(identical(colnames(mat), sample_meta$sample_id))
```

## 4. Baseline heatmap + palettes

```{r palettes, fig.width=9, fig.height=5}
mins <- min(mat, na.rm = TRUE)
maxs <- max(mat, na.rm = TRUE)
mids <- (mins + maxs) / 2
palette_blue_red <- circlize::colorRamp2(c(mins, mids, maxs), c('#0c2c84','#f7fbff','#b30000'))
palette_viridis <- circlize::colorRamp2(c(mins, mids, maxs), viridisLite::viridis(3))

ht_blue <- Heatmap(mat, name = 'value', col = palette_blue_red, na_col = na_color,
                   cluster_rows = FALSE, cluster_columns = FALSE,
                   show_row_names = FALSE, show_column_names = FALSE)
ht_viridis <- Heatmap(mat, name = 'value', col = palette_viridis, na_col = na_color,
                      cluster_rows = FALSE, cluster_columns = FALSE,
                      show_row_names = FALSE, show_column_names = FALSE)

ht_blue
ht_viridis
```

## 5. Annotations + ordering

```{r annotate, fig.width=10, fig.height=6}
sample_meta$post_ab <- ifelse(sample_meta$day == 0, 'baseline', 'post')
col_ann <- HeatmapAnnotation(
  treatment = sample_meta$treatment_group,
  status = sample_meta$post_ab,
  annotation_name_side = 'left'
)
order_idx <- order(sample_meta$mouse_id, sample_meta$day)
mat_ordered <- mat[, order_idx]
col_ann_ordered <- col_ann[order_idx]

Heatmap(
  mat_ordered,
  name = 'value',
  col = palette_blue_red,
  top_annotation = col_ann_ordered,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  na_col = na_color
)
```

## 6. Annotation enhancements

```{r annotate-advanced, fig.width=10, fig.height=6}
treatment_levels <- unique(sample_meta$treatment_group)
treatment_cols <- grDevices::hcl.colors(length(treatment_levels), palette = "Set2")
names(treatment_cols) <- treatment_levels
status_cols <- c(baseline = '#4daf4a', post = '#984ea3')
day_col_fun <- circlize::colorRamp2(range(sample_meta$day), c('#f7fbff','#084594'))
col_ann_rich <- HeatmapAnnotation(
  treatment = anno_simple(sample_meta$treatment_group, col = treatment_cols),
  status = anno_simple(sample_meta$post_ab, col = status_cols),
  day = anno_simple(sample_meta$day, col = day_col_fun),
  mouse = anno_text(sample_meta$mouse_id, rot = 90, gp = grid::gpar(fontsize = 6)),
  annotation_name_side = 'left'
)
col_ann_rich_ordered <- col_ann_rich[order_idx]
column_split <- factor(sample_meta$treatment_group[order_idx], levels = treatment_levels)
row_split <- factor(wide_df$Genome, levels = unique(wide_df$Genome))
row_titles <- sub('_.*', '', levels(row_split))

Heatmap(
  mat_ordered,
  name = 'value',
  col = palette_blue_red,
  top_annotation = col_ann_rich_ordered,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_split = column_split,
  show_row_names = FALSE,
  show_column_names = FALSE,
  na_col = na_color,
  row_split = row_split,
  row_title = row_titles,
  row_title_gp = grid::gpar(fontsize = 9)
)
```

## 7. Row variance filter

```{r variance, fig.width=10, fig.height=6}
row_var <- apply(mat, 1, var, na.rm = TRUE)
keep_idx <- order(row_var, decreasing = TRUE)[seq_len(min(100, nrow(mat)))]
mat_topvar <- mat[keep_idx, order_idx]
row_split_top <- droplevels(row_split[keep_idx])
Heatmap(
  mat_topvar,
  name = 'value',
  col = palette_blue_red,
  top_annotation = col_ann_rich_ordered,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_split = column_split,
  show_row_names = FALSE,
  show_column_names = FALSE,
  na_col = na_color,
  row_split = row_split_top,
  row_title_gp = grid::gpar(fontsize = 9)
)
```

## 8. Final heatmap + PDF export

```{r final, fig.width=10, fig.height=6}
ht_final <- Heatmap(
  mat_ordered,
  name = 'value',
  col = palette_blue_red,
  top_annotation = col_ann_rich_ordered,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_split = column_split,
  show_row_names = FALSE,
  show_column_names = FALSE,
  na_col = na_color,
  row_split = row_split,
  row_title = row_titles,
  row_title_gp = grid::gpar(fontsize = 9),
  column_title = 'All genomes (dataset3)'
)
draw(ht_final)
```

```{r export, eval=FALSE}
pdf(pdf_path, width = 11, height = 7)
draw(ht_final)
dev.off()
cat('Saved heatmap to', pdf_path, '\n')
```

## 9. Notes

- Palette choice: the blue-white-red ramp emphasizes deviations from mid values.
- Column ordering by mouse/day reveals antibiotic pulses more clearly than the
  CSV default.
- Variance filtering is helpful when presenting in class; it trims the figure to
  the most dynamic SNPs and speeds up PDF export.
- Extra annotations (day gradient + mouse labels + row splits) plus column
  splits by treatment mirror the annotation tutorial and keep the story tied to
  the experimental design.
