---
title: "Day 2 – 04 Full Heatmap (Exercises)"
author: "Seminar practice worksheet"
output:
  html_document:
    toc: true
    toc_depth: 2
---

This practice notebook guides you through rebuilding the "all genomes" heatmap
(`dataset3_subset*.csv`). Compared with earlier exercises, you now juggle a
larger matrix, multiple annotation layers, and custom ordering. Follow the
prompts, fill in each `# TODO`, and keep notes on the design decisions you make.

> **Tip:** Run `scripts/00_prepare.Rmd` first so `ComplexHeatmap`, `circlize`,
> and helper packages are ready. If the dataset is missing, regenerate it via
> `data/00_prepare_dataset.Rmd`.

## 1. Packages, paths, helper settings

Set yourself up for success by loading packages, pointing to the dataset files,
and defining constants you will re-use (e.g., the NA color and PDF output path).
Add any palettes you plan to try so they are available later.

```{r ex1-setup, eval=FALSE}
# TODO: library(ComplexHeatmap); library(circlize); library(viridisLite)
# TODO: subset_path <- file.path('..','data','dataset3_subset.csv')
#       long_path   <- file.path('..','data','dataset3_subset_long.csv')
#       pdf_path    <- file.path('..','pdf','04_full_heatmap_exercise.pdf')
# TODO: na_color <- '#dcdcdc'
# TODO: set.seed(...) if you plan to shuffle or sample
```

*Reflection:* Why do we keep both the wide and long versions even though the
heatmap consumes only the wide table?

## 2. Load data and report missingness

Read both tables, print their dimensions, and log the total number of NA values
in the `value` column. If NAs are present, produce a matrix (mouse vs day) that
shows where they occur so you can make informed filtering choices later.

```{r ex2-load, eval=FALSE}
# TODO: wide_df <- read.csv(subset_path, check.names = FALSE, stringsAsFactors = FALSE)
# TODO: long_df <- read.csv(long_path, check.names = FALSE, stringsAsFactors = FALSE)
# TODO: cat('Wide rows x cols:', nrow(wide_df), ncol(wide_df), '\n')
# TODO: cat('Long rows x cols:', nrow(long_df), ncol(long_df), '\n')
# TODO: na_total <- sum(is.na(long_df$value))
# TODO: if (na_total > 0) { build tapply(...) to locate them }
```

*Question:* Does the larger dataset introduce NA clusters concentrated in a
single mouse/treatment?

## 3. Build the numeric matrix

Construct the heatmap matrix using the sample columns, ensuring you keep row
labels informative for debugging. Reuse or rebuild a `sample_meta` data frame
that captures `mouse_id`, `day`, `treatment_group`, and `sample_id`. This
metadata will power your annotations and ordering.

```{r ex3-matrix, eval=FALSE}
# TODO: sample_cols <- setdiff(names(wide_df), c('Genome','snp_id','Position'))
# TODO: mat <- as.matrix(wide_df[, sample_cols]); mode(mat) <- 'numeric'
# TODO: rownames(mat) <- paste(wide_df$Genome, wide_df$snp_id, sep = ' | ')
# TODO: sample_meta <- unique(long_df[, c('mouse_id','day','treatment_group')])
# TODO: sample_meta$sample_id <- paste(sample_meta$mouse_id, sample_meta$day, sep='-')
# TODO: sample_meta <- sample_meta[match(colnames(mat), sample_meta$sample_id), ]
```

*Check:* Use `stopifnot(identical(colnames(mat), sample_meta$sample_id))` to
catch mismatches early.

## 4. Baseline heatmap and color experiments

Start with a minimal `Heatmap` (no clustering, no annotations) so you can verify
the matrix orientation. Next, experiment with at least two color palettes by
constructing separate `colorRamp2` functions. Document which palette you prefer
and why (contrast, perceptual ordering, etc.).

```{r ex4-colors, eval=FALSE}
# TODO: mins <- min(mat, na.rm = TRUE); maxs <- max(mat, na.rm = TRUE); mids <- (mins+maxs)/2
# TODO: palette_a <- circlize::colorRamp2(c(mins, mids, maxs), c('#0c2c84','#f7fbff','#b30000'))
# TODO: palette_b <- circlize::colorRamp2(c(mins, mids, maxs), viridisLite::viridis(3))
# TODO: Heatmap(mat, name='value', col=palette_a, na_col = na_color)
# TODO: Heatmap(mat, name='value', col=palette_b, na_col = na_color)
```

*Reflection:* Which palette makes low values easiest to distinguish once
annotations are added?

## 5. Column annotations and ordering

Build annotations that explain treatment group and baseline/post-antibiotic
status. Then define an ordering (e.g., order by `mouse_id`, then `day`) and
apply it consistently to both the matrix and the annotation object. Draw the
heatmap again to confirm columns appear in the intended sequence.

```{r ex5-annotate, eval=FALSE}
# TODO: sample_meta$post_ab <- ifelse(sample_meta$day == 0, 'baseline', 'post')
# TODO: col_ann <- HeatmapAnnotation(
#         treatment = sample_meta$treatment_group,
#         status = sample_meta$post_ab,
#         annotation_name_side = 'left'
#       )
# TODO: order_idx <- order(sample_meta$mouse_id, sample_meta$day)
# TODO: mat_ordered <- mat[, order_idx]
# TODO: col_ann_ordered <- col_ann[order_idx]
# TODO: Heatmap(mat_ordered, name='value', top_annotation = col_ann_ordered,
#               cluster_rows = FALSE, cluster_columns = FALSE, na_col = na_color)
```

*Check:* Are replicates or treatment switches now easier to spot?

## 6. Annotation enhancements, column splits, and readable row titles

Recreate the richer annotation stack from `03_heatmap_annotations.Rmd`, but now
apply it to the full dataset. Ideas to try:

- Encode `treatment_group` with a named color vector so the legend matches your
  slide deck.
- Add a continuous day gradient using `anno_simple()` so viewers can track the
  time axis directly in the annotation bar.
- Overlay mouse IDs (rotated text or a thin color strip) to highlight replicate
  structure.
- Optionally, split rows by genome (`row_split`) to mimic the multi-panel layout
  from the guided notebook, and either shorten those labels or shrink the font
  so they do not overlap in the PDF.
- Add a `column_split` (e.g., by treatment group) so the plot is chunked into
  digestible vertical blocks in addition to the chronological ordering.

```{r ex6-annotations, eval=FALSE}
# TODO: treatment_cols <- c(...); status_cols <- c(...)
# TODO: day_col_fun <- circlize::colorRamp2(range(sample_meta$day), c('#f7fbff','#084594'))
# TODO: col_ann_rich <- HeatmapAnnotation(
#         treatment = anno_simple(sample_meta$treatment_group, col = treatment_cols),
#         status    = anno_simple(sample_meta$post_ab, col = status_cols),
#         day       = anno_simple(sample_meta$day, col = day_col_fun),
#         mouse     = anno_text(sample_meta$mouse_id, rot = 90, gp = grid::gpar(fontsize = 6)),
#         annotation_name_side = 'left'
#       )
# TODO: col_ann_rich <- col_ann_rich[order_idx]
# TODO: column_split <- factor(sample_meta$treatment_group[order_idx], levels = treatment_levels)
# TODO: row_split <- factor(wide_df$Genome, levels = unique(wide_df$Genome))
# TODO: row_titles <- gsub('_.*', '', levels(row_split)) # or set fontsize via row_title_gp
# TODO: Heatmap(mat_ordered, name='value', col = palette_a, top_annotation = col_ann_rich,
#               cluster_rows = TRUE, row_split = row_split, column_split = column_split,
#               na_col = na_color, show_row_names = FALSE, show_column_names = FALSE,
#               row_title = row_titles, row_title_gp = grid::gpar(fontsize = 9))
```

*Reflection:* Which annotation layer (treatment colors, day gradient, mouse IDs)
helped the most when interpreting the plot? Keep notes for your presentation.

## 7. Row filtering via variance

Large matrices can hide structure. Compute per-row variance, keep the top 100
rows (or another threshold), and draw a diagnostic heatmap to see whether the
higher-variance SNPs highlight patterns that were previously buried. Remember to
reapply the column ordering and annotations.

```{r ex7-variance, eval=FALSE}
# TODO: row_var <- apply(mat, 1, var, na.rm = TRUE)
# TODO: keep_idx <- order(row_var, decreasing = TRUE)[seq_len(min(100, nrow(mat)))]
# TODO: mat_topvar <- mat[keep_idx, order_idx]
# TODO: Heatmap(mat_topvar, name='value', top_annotation = col_ann_rich,
#               cluster_rows = TRUE, cluster_columns = FALSE, column_split = column_split,
#               na_col = na_color, row_split = row_split[keep_idx], show_row_names = FALSE,
#               row_title_gp = grid::gpar(fontsize = 9))
```

*Reflection:* Does variance filtering change the biological story you would tell?

## 8. Final polish and PDF export

Combine your favorite palette, annotations, column order, and a row clustering
strategy into a final heatmap. Add a column title and tweak legend names if
needed. Export the final figure to `pdf_path` so it can be shared outside the
notebook.

```{r ex8-final, eval=FALSE}
# TODO: ht_final <- Heatmap(
#         mat_ordered,
#         name = 'value',
#         col = palette_a,
#         top_annotation = col_ann_rich,
#         cluster_rows = TRUE,
#         cluster_columns = FALSE,
#         column_split = column_split,
#         show_row_names = FALSE,
#         show_column_names = FALSE,
#         na_col = na_color,
#         row_split = row_split,
#         row_title = row_titles,
#         row_title_gp = grid::gpar(fontsize = 9),
#         column_title = 'All genomes (dataset3)'
#       )
# TODO: draw(ht_final)
# TODO: pdf(pdf_path, width = 11, height = 7); draw(ht_final); dev.off();
#       cat('Saved heatmap to', pdf_path, '\n')
```

*Question:* Which design choices would you highlight if you presented this plot
in lab meeting (ordering rationale, annotation selection, etc.)?

## 9. Notes and extensions

Use this space to capture observations, alternative palettes, or next steps
(e.g., adding row annotations, trying `row_split`, exporting PNGs).

```{r ex8-notes, eval=FALSE}
# TODO: jot down thoughts, ideas, or additional code experiments
```

Once you're happy with the exercise output, compare it against the solution
notebook (`scripts/04_full_heatmap_exercises_solution.Rmd`) to fill any gaps.
